---
title: "Evaluating Vegetation and Groundwater in Southern Africa"
author: "Louis Quigley"
institute: "University of Illinois Chicago"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r load-packages, include=FALSE}
library(tidyverse)
library(knitr)
library(lubridate)
library(broom)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.retina = 3,
  dpi = 300,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "80%",
  message = FALSE,
  warning = FALSE
)
```

```{r load-data, include=F}
# Load data
data <- read_csv("SVI_SGI_timeseries.csv", show_col_types = FALSE)

# Convert date from YYYYMM format to a real date
data <- data %>%
  mutate(
    date = as.character(date),
    year = as.numeric(substr(date, 1, 4)),
    month = as.numeric(substr(date, 5, 6)),
    date = ymd(paste(year, month, "01", sep = "-"))
  ) %>%
  arrange(date) %>%
  drop_na(SVI, SGI)

# Check data
glimpse(data)
summary(data)
```

```{r confidence-intervals, include=F}
# Bootstrap confidence intervals for mean SVI and SGI

set.seed(480)
B <- 1000

boot_ci_mean <- function(x, B = 1000) {
  x <- x[!is.na(x)]
  
  boot_means <- replicate(
    B,
    mean(sample(x, size = length(x), replace = TRUE))
  )
  
  ci <- quantile(boot_means, probs = c(0.025, 0.975))
  
  list(
    n = length(x),
    mean = mean(x),
    sd = sd(x),
    se = sd(x) / sqrt(length(x)),
    ci_lower = ci[1],
    ci_upper = ci[2],
    boot_means = boot_means
  )
}

# Run bootstrap for each index
svi_boot <- boot_ci_mean(data$SVI, B = B)
sgi_boot <- boot_ci_mean(data$SGI, B = B)

# Table for display
ci_display <- tibble(
  Index = c("SVI", "SGI"),
  n = c(svi_boot$n, sgi_boot$n),
  `Mean anomaly` = c(svi_boot$mean, sgi_boot$mean),
  SD = c(svi_boot$sd, sgi_boot$sd),
  SE = c(svi_boot$se, sgi_boot$se),
  `95% CI lower` = c(svi_boot$ci_lower, sgi_boot$ci_lower),
  `95% CI upper` = c(svi_boot$ci_upper, sgi_boot$ci_upper)
) %>%
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

```{r t-tests, include=FALSE}
#two-sample t-tests: 1982-1999 vs. 2000-2024

data_period <- data %>%
  mutate(
    period = case_when(
      year >= 1982 & year <= 1999 ~ "1982-1999",
      year >= 2000 & year <= 2024 ~ "2000-2024",
      TRUE ~ NA_character_
    ),
    period = factor(period, levels = c("1982-1999", "2000-2024"))
  ) %>%
  drop_na(period)

# SVI two-sample t-test

svi_x1 <- data_period %>%
  filter(period == "1982-1999") %>%
  pull(SVI)

svi_x2 <- data_period %>%
  filter(period == "2000-2024") %>%
  pull(SVI)

svi_xbar1 <- mean(svi_x1, na.rm = TRUE)
svi_xbar2 <- mean(svi_x2, na.rm = TRUE)

svi_s1 <- sd(svi_x1, na.rm = TRUE)
svi_s2 <- sd(svi_x2, na.rm = TRUE)

svi_n1 <- sum(!is.na(svi_x1))
svi_n2 <- sum(!is.na(svi_x2))

svi_mean_diff <- svi_xbar1 - svi_xbar2

svi_se_diff <- sqrt((svi_s1^2 / svi_n1) + (svi_s2^2 / svi_n2))

svi_t_stat <- svi_mean_diff / svi_se_diff

svi_df_welch <- ((svi_s1^2 / svi_n1 + svi_s2^2 / svi_n2)^2) /
  (((svi_s1^2 / svi_n1)^2 / (svi_n1 - 1)) +
     ((svi_s2^2 / svi_n2)^2 / (svi_n2 - 1)))

svi_p_value <- 2 * pt(-abs(svi_t_stat), df = svi_df_welch)

svi_ttest_manual <- tibble(
  index = "SVI",
  mean_1982_1999 = svi_xbar1,
  mean_2000_2024 = svi_xbar2,
  mean_diff = svi_mean_diff,
  se_diff = svi_se_diff,
  t_stat = svi_t_stat,
  df_welch = svi_df_welch,
  p_value = svi_p_value
)

# Compare with automated t.test()
svi_ttest_auto <- t.test(svi_x1, svi_x2)

# SGI two-sample t-test

sgi_x1 <- data_period %>%
  filter(period == "1982-1999") %>%
  pull(SGI)

sgi_x2 <- data_period %>%
  filter(period == "2000-2024") %>%
  pull(SGI)

sgi_xbar1 <- mean(sgi_x1, na.rm = TRUE)
sgi_xbar2 <- mean(sgi_x2, na.rm = TRUE)

sgi_s1 <- sd(sgi_x1, na.rm = TRUE)
sgi_s2 <- sd(sgi_x2, na.rm = TRUE)

sgi_n1 <- sum(!is.na(sgi_x1))
sgi_n2 <- sum(!is.na(sgi_x2))

sgi_mean_diff <- sgi_xbar1 - sgi_xbar2

sgi_se_diff <- sqrt((sgi_s1^2 / sgi_n1) + (sgi_s2^2 / sgi_n2))

sgi_t_stat <- sgi_mean_diff / sgi_se_diff

sgi_df_welch <- ((sgi_s1^2 / sgi_n1 + sgi_s2^2 / sgi_n2)^2) /
  (((sgi_s1^2 / sgi_n1)^2 / (sgi_n1 - 1)) +
     ((sgi_s2^2 / sgi_n2)^2 / (sgi_n2 - 1)))

sgi_p_value <- 2 * pt(-abs(sgi_t_stat), df = sgi_df_welch)

sgi_ttest_manual <- tibble(
  index = "SGI",
  mean_1982_1999 = sgi_xbar1,
  mean_2000_2024 = sgi_xbar2,
  mean_diff = sgi_mean_diff,
  se_diff = sgi_se_diff,
  t_stat = sgi_t_stat,
  df_welch = sgi_df_welch,
  p_value = sgi_p_value
)

# compare with t.test()
sgi_ttest_auto <- t.test(sgi_x1, sgi_x2)

# display table for slides

ttest_display <- bind_rows(
  svi_ttest_manual,
  sgi_ttest_manual
) %>%
  mutate(
    across(
      c(mean_1982_1999, mean_2000_2024, mean_diff, se_diff, t_stat, df_welch),
      \(x) round(x, 3)
    ),
    p_value = format.pval(p_value, digits = 3)
  ) %>%
  rename(
    `Index` = index,
    `Mean 1982-1999` = mean_1982_1999,
    `Mean 2000-2024` = mean_2000_2024,
    `Mean difference` = mean_diff,
    `SE difference` = se_diff,
    `t statistic` = t_stat,
    `df` = df_welch,
    `p-value` = p_value
  )
```

```{r same-month-linear-regression, include=FALSE}
# Same-month linear regression

df_lm <- data %>%
  select(date, SVI, SGI) %>%
  drop_na(SVI, SGI)

lm_same_month <- lm(SGI ~ SVI, data = df_lm)

same_month_stats <- tidy(lm_same_month) %>%
  mutate(across(where(is.numeric), \(x) round(x, 3)))

same_month_fit <- glance(lm_same_month)

same_month_slope <- tidy(lm_same_month) %>%
  filter(term == "SVI")

same_month_p <- same_month_slope$p.value
```

```{r lagged-regression, include=FALSE}
# Lagged linear regression: -12 to +12 months

run_lagged_model <- function(lag_months, data) {
  
  if (lag_months > 0) {
    lagged_data <- data %>%
      mutate(SGI_shifted = lead(SGI, lag_months))
  }
  
  if (lag_months < 0) {
    lagged_data <- data %>%
      mutate(SGI_shifted = lag(SGI, abs(lag_months)))
  }
  
  if (lag_months == 0) {
    lagged_data <- data %>%
      mutate(SGI_shifted = SGI)
  }
  
  lagged_data <- lagged_data %>%
    drop_na(SVI, SGI_shifted)
  
  model <- lm(SGI_shifted ~ SVI, data = lagged_data)
  
  tidy(model) %>%
    filter(term == "SVI") %>%
    mutate(
      lag_months = lag_months,
      r_squared = glance(model)$r.squared,
      rmse = glance(model)$sigma,
      n = nrow(lagged_data)
    )
}

lag_results <- map_dfr(
  -12:12,
  ~ run_lagged_model(lag_months = .x, data = data)
)

best_lag <- lag_results %>%
  slice_max(r_squared, n = 1, with_ties = FALSE)

best_lag_value <- best_lag$lag_months[1]

best_lag_display <- best_lag %>%
  transmute(
    `Best lag` = lag_months,
    `R²` = round(r_squared, 3),
    `RMSE` = round(rmse, 3),
    `n` = n
  )
```

# Research question

.pull-left[
## How are vegetation and groundwater drought anomalies related in Southern Africa?

**Sub-questions**

- Are the means of standardized indices centered near normal?
- Did average conditions change between 1982–1999 and 2000–2024?
- Is there a lagged relationship between SVI and SGI?
]

.pull-right[
.center[
**Significance**

Vegetation and groundwater are important environmental resources. Both NDVI and water table depth are climate-sensitive, but they may respond to drought on different timescales.

Understanding their relationship can help evaluate compound drought impacts.
]
]

---

# Data

**Monthly time series of standardized NDVI and water table depths (WTD) from 1982–2024**  
**515 total observations**

| System | Variable | Standardized Variable |
|---|---|---|
| Vegetation | NDVI | Standardized Vegetation Index (SVI) |
| Groundwater | Modeled WTD | Standardized Groundwater Index (SGI) |

--

Standardization was done via a non-parametric normal scores transformation:

- mean becomes 0
- expresses variables as number of SD above or below the mean

<br>

NDVI -> Standardized Vegetation Index (SVI)
WTD -> Standardized Groundwater Index (SGI)

---

# Statistical methods

| Question | Method |
|---|---|
| What is the mean anomaly across 1982–2024? | Bootstrap 95% confidence interval | 
| Did means change after 2000? | Two-sample t-test |
| Is SVI linearly related to SGI? | Linear regression | 

-The lagged analysis repeats the same linear regression after shifting SGI from SVI by -12 to +12 months. 

---

# Mean anomaly estimates
.pull-left[
```{r svi-bootstrap-plot, echo=FALSE, fig.width=4.5, fig.asp=0.8, out.width="100%"}
ggplot(tibble(boot_mean = svi_boot$boot_means), aes(x = boot_mean)) +
  geom_histogram(bins = 40, alpha = 0.8) +
  geom_vline(
    xintercept = c(svi_boot$ci_lower, svi_boot$ci_upper),
    color = "red",
    linewidth = 1
  ) +
  geom_vline(
    xintercept = svi_boot$mean,
    linetype = "dashed"
  ) +
  theme_classic(base_size = 14) +
  labs(
    x = "Bootstrap mean SVI",
    y = "Count",
    title = "SVI"
  )
```
]

.pull-right[
```{r sgi-bootstrap-plot, echo=FALSE, fig.width=4.5, fig.asp=0.8, out.width="100%"}
  ggplot(tibble(boot_mean = sgi_boot$boot_means), aes(x = boot_mean)) +
  geom_histogram(bins = 40, alpha = 0.8) +
  geom_vline(
    xintercept = c(sgi_boot$ci_lower, sgi_boot$ci_upper),
    color = "red",
    linewidth = 1
  ) +
  geom_vline(
    xintercept = sgi_boot$mean,
    linetype = "dashed"
  ) +
  theme_classic(base_size = 14) +
  labs(
    x = "Bootstrap mean SGI",
    y = "Count",
    title = "SGI"
  )
```
]

.footnote[
The black dashed line represents the bootstrap mean. Red vertical lines show the bootstrap 95% confidence interval.
]

---

# Mean anomaly estimates: summary table

```{r ci-table, echo=FALSE}
kable(
  ci_display,
  format = "html",
  caption = "Bootstrap 95% confidence intervals for mean SVI and SGI"
)
```

---

# Did average conditions change after 1999?
  
```{r ttest-table, echo=FALSE}
kable(
  ttest_display,
  format = "html",
  caption = "two-sample t-test: 1982–1999 vs. 2000–2024"
)
```
--

**Interpretation**

- t-test evaluates whether the difference in the period means is equal to zero (null-hypothesis)
- Mean SVI was significantly higher in 1982–1999 than in 2000–2024
- Mean SGI did not differ significantly between periods

---
# Visualizing the two-period comparison

.pull-left[
```{r svi-period-boxplot, echo=FALSE, fig.width=4.5, fig.asp=0.8, out.width="100%"}
ggplot(data_period, aes(x = period, y = SVI, fill = period)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = c("1982-1999" = "forestgreen",
  "2000-2024"="darkorange"
  )) +
  theme_classic(base_size = 14) +
  labs(
    x = NULL,
    y = "SVI",
    title = "SVI by period"
  ) +
  theme(legend.position = "none")
```
]

.pull-right[
```{r sgi-period-boxplot, echo=FALSE, fig.width=4.5, fig.asp=0.8, out.width="100%"}
ggplot(data_period, aes(x = period, y = SGI, fill = period)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = c(
    "1982-1999" = "steelblue",
    "2000-2024" = "darkorange"
  ))+
  theme_classic(base_size = 14) +
  labs(
    x = NULL,
    y = "SGI",
    title = "SGI by period"
  ) +
  theme(legend.position = "none")
```
]
---

# Concurrent SVI–SGI relationship

.pull-left[
```{r same-month-scatter-smooth, echo=FALSE, fig.width=4.5, fig.asp=0.8, out.width="100%"}
ggplot(df_lm, aes(x = SVI, y = SGI)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_classic(base_size = 14) +
  labs(
    x = "SVI",
    y = "SGI",
    title = "Scatterplot with linear fit"
  )
```
]

.pull-right[
| Statistic | Value |
|---|---:|
| R² | `r round(same_month_fit$r.squared, digits = 4)` |

<br>
SVI explains `r round(same_month_fit$r.squared * 100, 2)`% of the same-month variation in SGI.
<br>

The same-month relationship is very weak :(
<br>
--
But that's ok!
]
---

# Does the relationship depend on timing?
Linear regression repeated on shifted SGI and SVI
<br>

Shifted SGI -12 to +12 months

<br>

--
positive lag means SVI leads SGI
<br>
Negative lag means SVI follows SGI
---

# Lagged SVI–SGI model fit

.pull-left[
```{r lag-r2-plot, echo=FALSE, fig.width=4.8, fig.asp=0.8, out.width="100%"}
ggplot(lag_results, aes(x = lag_months, y = r_squared)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = best_lag_value, linetype = "dashed") +
  theme_classic(base_size = 14) +
  labs(
    x = "Lag in months\nPositive = SVI leads SGI",
    y = expression(R^2),
    title = expression(R^2 ~ "across monthly lags")
  )
```
]
.pull-right[
| Statistic | Value |
|---|---:|
| Best lag | `r best_lag_value` months |
| R² | `r round(best_lag$r_squared, digits = 4)` |

Changes in SVI most likely precedes SGI by two months
<br>

Low R² could be due to averaging a large area
]

---

# Key takeaways

--

1. Distribution of SVI and SGI are centered close to their long-term mean

--

2. Mean SVI was higher in 1982–1999 than in 2000–202

--

3. Mean SGI did not differ significantly between periods.

--

4. The relationship peaked at `r best_lag_value` months, but R² was low